---
title: "Visited random points and transects"
format: html
---

## Overview

Field crews have started revisiting the randomly-placed ground-truth
points/transects and recording presence data (`Prsnc_M`, `Prsnc_J`, `Prsnc_S`
- one column per visit). A point is treated as *visited* once any of those
three columns is non-`NA`, regardless of the value recorded (0 or a count).

Goal here is just to pull two subsets out of `ground_truth_median-VISITED.gpkg`:

- **Visited points** - the random `base_pts`/`os_pts` that have been visited.
- **Visited transects** - the *entire* transect (all points along it, not
  just the visited ones) for any transect (`transects_known`/`transects_other`,
  grouped by `patch_ID`) that has at least one visited point.

```{r}
library(sf)
library(dplyr)
library(tidyverse)
```

```{r import}
gt_path <- file.path('..', '..', 'data', 'GroundTruthPts', 'ground_truth_median-VISITED.gpkg')

base_pts <- st_read(gt_path, layer = 'base_pts', quiet = TRUE)
os_pts <- st_read(gt_path, layer = 'os_pts', quiet = TRUE)
transects_known <- st_read(gt_path, layer = 'transects_known', quiet = TRUE)
transects_other <- st_read(gt_path, layer = 'transects_other', quiet = TRUE)
```

## Visited points

```{r visited points}
was_visited <- function(df) {
  df |> mutate(visited = if_any(c(Prsnc_M, Prsnc_J, Prsnc_S), ~ !is.na(.)))
}

points <- bind_rows(
  was_visited(base_pts) |> mutate(source = 'base_pts'),
  was_visited(os_pts) |> mutate(source = 'os_pts')
)

visited_points <- points |> filter(visited)
nrow(visited_points)
```

## Visited transects

```{r visited transects}
transects <- bind_rows(
  was_visited(transects_known) |> mutate(source = 'transects_known'),
  was_visited(transects_other) |> mutate(source = 'transects_other')
)

visited_patch_ids <- transects |>
  st_drop_geometry() |>
  filter(visited) |>
  distinct(source, patch_ID)

visited_transects <- transects |>
  semi_join(visited_patch_ids, by = c('source', 'patch_ID'))

n_distinct(paste(visited_transects$source, visited_transects$patch_ID))
nrow(visited_transects)
```

## Quick check

```{r}
ggplot() +
  geom_sf(data = visited_transects, aes(color = source)) +
  geom_sf(data = visited_points, aes(color = source), size = 2) +
  theme_void()
```

## Write out

```{r write}
out_path <- file.path('..', '..', 'data', 'GroundTruthPts', 'visited_subset.gpkg')

st_write(visited_points, out_path, layer = 'visited_points', append = FALSE, quiet = TRUE)
st_write(visited_transects, out_path, layer = 'visited_transects', append = FALSE, quiet = TRUE)
```
