---
title: "SDManuscript maps"
format: html
---

Builds the manuscript's map figures and writes each composite to `figures/`
as a PNG. Kept separate from SDManuscript.Rmd because the final rasters are
large and slow to crop/mask/render - knitting the manuscript itself
shouldn't have to pay that cost every time. Render this file whenever the
source rasters change; SDManuscript.Rmd just pulls in whatever PNGs it
finds in `figures/` via `knitr::include_graphics()`.

```{r setup}
library(ggplot2)
library(patchwork)
library(sf)
library(terra)
source('theme_manuscript.R')

fig_dir <- 'figures'
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

## Final suitability and density maps


```{r zoom bboxes}
## Manually chosen zoom windows for the density figure's inset panels, read
## off a basemap in WGS84 lon/lat - each entry is NW corner then SE corner
## (lat, lon). Kept in WGS84 here; reprojected to the raster's CRS wherever
## they're actually used to crop (see `zoom_bbs` below).
zoom_corners <- list(
  cb     = c(nw_lat = 39.015644, nw_lon = -107.034565, se_lat = 39.007642, se_lon = -107.0207323),
  coche  = c(nw_lat = 38.177820, nw_lon = -106.705831, se_lat = 38.174130, se_lon = -106.696740),
  antero = c(nw_lat = 38.984090, nw_lon = -105.993880, se_lat = 38.974038, se_lon = -105.976990)
)

zoom_bbs_ll <- lapply(zoom_corners, function(z) {
  sf::st_bbox(
    c(xmin = z[['nw_lon']], xmax = z[['se_lon']], ymin = z[['se_lat']], ymax = z[['nw_lat']]),
    crs = 4326
  )
})
```


```{r Final Suitability and Density Maps, eval = FALSE}
## TODO: paths below point to the final refit outputs (2.8) - not yet
## produced (that model uses ground-truth data still being collected) -
## so this chunk stays eval=FALSE until they exist.

bbs_sf <- sf::st_read(
  file.path('..', '..', 'data', 'Cartography', 'bboxes_for_manuscript.gpkg'),
  quiet = TRUE
)
region_names <- c('cb', 'coche', 'antero')
bbs <- setNames(lapply(1:3, function(i) sf::st_bbox(bbs_sf[i, ])), region_names)

suit_r <- terra::rast(file.path('..', '..', 'results', 'suitability_maps', 'final-Pr.tif'))
aoa_r  <- terra::rast(file.path('..', '..', 'results', 'suitability_maps', 'final-AOA.tif'))
dens_r <- terra::rast(file.path('..', '..', 'results', 'count_models', 'final-density.tif'))

## AOA == 0 pixels are outside the model's area of applicability - not real
## predictions - so null them out, and speckle that area on the plot rather
## than leaving it a blank gap.
suit_r  <- mask_aoa(suit_r, aoa_r)
speck   <- aoa_speckle(aoa_r)

## ---- final habitat suitability: full domain (left, full height) + the
## three focal regions to the right, each panel the same size so smaller
## regions read at a visibly larger scale than the full domain ----
suit_full   <- mapPanel(
  rasterToDF(suit_r),
  scale_fill_suitability, 
  'Full domain', 
  bboxes = bbs, 
  speckle = speck
  )
suit_cb     <- mapPanel(rasterToDF(suit_r, bbs$cb),     scale_fill_suitability, 'Crested Butte', speckle = speck)
suit_coche  <- mapPanel(rasterToDF(suit_r, bbs$coche),  scale_fill_suitability, 'Cochetopa', speckle = speck)
suit_antero <- mapPanel(rasterToDF(suit_r, bbs$antero), scale_fill_suitability, 'Antero', speckle = speck)

p_suit_final <- (suit_full + (suit_cb / suit_coche / suit_antero) +
  plot_layout(widths = c(2, 1), guides = 'collect')) +
  plot_annotation(tag_levels = 'A') &
  theme_manuscript_tags() &
  theme(legend.position = 'bottom')
p_suit_final
ggsave(file.path(fig_dir, 'suitability_final.png'), p_suit_final, width = 9, height = 7, dpi = 300)
```

```{r Final Density Maps, eval = FALSE}
## ---- final census/density surfaces: one row per site, stacked
## vertically (no full-domain panel here - density is only modelled/
## meaningful within the surveyed regions). Each row pairs the site-level
## map (with a nested zoom bbox outlined) against a further zoomed-in crop
## of that nested box. All sites share one fill scale/legend so density is
## comparable across sites, not just within one. ----
## Zoom windows are the manually chosen boxes from `zoom_bbs_ll` (WGS84),
## reprojected to the density raster's CRS.
dens_nested <- lapply(zoom_bbs_ll[names(bbs)], function(z) {
  sf::st_bbox(sf::st_transform(sf::st_as_sfc(z), terra::crs(dens_r)))
})

dens_df_cb     <- rasterToDF(dens_r, bbs$cb)
dens_df_coche  <- rasterToDF(dens_r, bbs$coche)
dens_df_antero <- rasterToDF(dens_r, bbs$antero)
dens_limits <- range(c(dens_df_cb$value, dens_df_coche$value, dens_df_antero$value), na.rm = TRUE)
dens_scale <- function() scale_fill_density(limits = dens_limits)

dens_cb_main     <- mapPanel(dens_df_cb,     dens_scale, 'Crested Butte', bboxes = list(zoom = dens_nested$cb))
dens_cb_zoom     <- mapPanel(rasterToDF(dens_r, dens_nested$cb),     dens_scale, 'Crested Butte (zoom)')
dens_coche_main  <- mapPanel(dens_df_coche,  dens_scale, 'Cochetopa', bboxes = list(zoom = dens_nested$coche))
dens_coche_zoom  <- mapPanel(rasterToDF(dens_r, dens_nested$coche),  dens_scale, 'Cochetopa (zoom)')
dens_antero_main <- mapPanel(dens_df_antero, dens_scale, 'Antero', bboxes = list(zoom = dens_nested$antero))
dens_antero_zoom <- mapPanel(rasterToDF(dens_r, dens_nested$antero), dens_scale, 'Antero (zoom)')

p_dens_final <- (
  (dens_cb_main | dens_cb_zoom) /
  (dens_coche_main | dens_coche_zoom) /
  (dens_antero_main | dens_antero_zoom) +
  plot_layout(guides = 'collect')) +
  plot_annotation(tag_levels = 'A') &
  theme_manuscript_tags() &
  theme(legend.position = 'bottom', legend.justification = 'center')
p_dens_final
ggsave(file.path(fig_dir, 'density_final.png'), p_dens_final, width = 8, height = 10, dpi = 300)
```

## Theme preview (real test raster, no final refit yet)

```{r Example Maps Theme Preview, fig.height = 6}
## Preview of theme_manuscript_map() styling using a real single-run
## suitability/AOA/SE triple dropped in results/suitability_maps for this
## purpose (not the final refit - that's still above, eval=FALSE until it
## exists). Delete this chunk once those are back to eval=TRUE.
## No real density test raster yet, so that panel stays on example_surface().
demo_fp <- file.path('..', '..', 'results', 'suitability_maps')
suit_r_demo <- terra::rast(file.path(demo_fp, '3arc-Iteration1-PA1:DO:0-Seed1017-Pr.tif'))
aoa_r_demo  <- terra::rast(file.path(demo_fp, '3arc-Iteration1-PA1:DO:0-Seed1017-AOA.tif'))
se_r_demo   <- terra::rast(file.path(demo_fp, '3arc-Iteration1-PA1:DO:0-Seed1017-SE.tif'))

## AOA == 0 pixels are outside the model's area of applicability - not real
## predictions - so null them out before plotting.
suit_r_demo <- mask_aoa(suit_r_demo, aoa_r_demo)
se_r_demo   <- mask_aoa(se_r_demo, aoa_r_demo)
demo_crs    <- terra::crs(suit_r_demo)
speck_demo  <- aoa_speckle(aoa_r_demo)

bbs_sf_demo <- sf::st_transform(
  sf::st_read(file.path('..', '..', 'data', 'Cartography', 'bboxes_for_manuscript.gpkg'), quiet = TRUE),
  demo_crs
)
region_names_demo <- c('cb', 'coche', 'antero')
bbs_demo <- setNames(lapply(1:3, function(i) sf::st_bbox(bbs_sf_demo[i, ])), region_names_demo)
dens_r_demo <- example_surface(sf::st_bbox(bbs_sf_demo), res = 100, range = c(0, 40), seed = 28)

## ---- suitability: full domain (left, north arrow since it's the
## overview panel) + the three focal regions to the right ----
suit_full_demo   <- mapPanel(rasterToDF(suit_r_demo), scale_fill_suitability,
                              'Full domain', crs = demo_crs, bboxes = bbs_demo, speckle = speck_demo, north_arrow = TRUE)
suit_cb_demo     <- mapPanel(rasterToDF(suit_r_demo, bbs_demo$cb),     scale_fill_suitability, 'Crested Butte', crs = demo_crs, speckle = speck_demo)
suit_coche_demo  <- mapPanel(rasterToDF(suit_r_demo, bbs_demo$coche),  scale_fill_suitability, 'Cochetopa', crs = demo_crs, speckle = speck_demo)
suit_antero_demo <- mapPanel(rasterToDF(suit_r_demo, bbs_demo$antero), scale_fill_suitability, 'Antero', crs = demo_crs, speckle = speck_demo)

p_suit_preview <- (suit_full_demo + (suit_cb_demo / suit_coche_demo / suit_antero_demo) +
  plot_layout(widths = c(2, 1), guides = 'collect')) +
  plot_annotation(
    tag_levels = 'A',
    title = 'Final ENM for Colorado Buckwheat') &
  theme_manuscript_tags() &
  theme(legend.position = 'bottom')
p_suit_preview
ggsave(file.path(fig_dir, 'suitability_preview.png'), p_suit_preview, width = 9, height = 7, dpi = 300)

## ---- density: one row per site, stacked vertically, each row pairing
## the site-level map (nested zoom bbox outlined) with a further zoomed-in
## crop of that nested box. All sites share one fill scale/legend. ----
dens_nested_demo <- lapply(bbs_demo, shrink_bbox, factor = 0.35)

dens_df_cb_demo     <- rasterToDF(dens_r_demo, bbs_demo$cb)
dens_df_coche_demo  <- rasterToDF(dens_r_demo, bbs_demo$coche)
dens_df_antero_demo <- rasterToDF(dens_r_demo, bbs_demo$antero)
dens_limits_demo <- range(c(dens_df_cb_demo$value, dens_df_coche_demo$value, dens_df_antero_demo$value), na.rm = TRUE)
dens_scale_demo <- function() scale_fill_density(limits = dens_limits_demo)

dens_cb_main_demo     <- mapPanel(dens_df_cb_demo,     dens_scale_demo, 'Crested Butte', crs = demo_crs, bboxes = list(zoom = dens_nested_demo$cb))
dens_cb_zoom_demo     <- mapPanel(rasterToDF(dens_r_demo, dens_nested_demo$cb),     dens_scale_demo, 'Crested Butte (zoom)', crs = demo_crs)
dens_coche_main_demo  <- mapPanel(dens_df_coche_demo,  dens_scale_demo, 'Cochetopa', crs = demo_crs, bboxes = list(zoom = dens_nested_demo$coche))
dens_coche_zoom_demo  <- mapPanel(rasterToDF(dens_r_demo, dens_nested_demo$coche),  dens_scale_demo, 'Cochetopa (zoom)', crs = demo_crs)
dens_antero_main_demo <- mapPanel(dens_df_antero_demo, dens_scale_demo, 'Antero', crs = demo_crs, bboxes = list(zoom = dens_nested_demo$antero))
dens_antero_zoom_demo <- mapPanel(rasterToDF(dens_r_demo, dens_nested_demo$antero), dens_scale_demo, 'Antero (zoom)', crs = demo_crs)

p_dens_preview <- (
  (dens_cb_main_demo | dens_cb_zoom_demo) /
  (dens_coche_main_demo | dens_coche_zoom_demo) /
  (dens_antero_main_demo | dens_antero_zoom_demo) +
  plot_layout(guides = 'collect')) +
  plot_annotation(tag_levels = 'A') &
  theme_manuscript_tags() &
  theme(legend.position = 'bottom', legend.justification = 'center')
p_dens_preview
ggsave(file.path(fig_dir, 'density_preview.png'), p_dens_preview, width = 8, height = 10, dpi = 300)
```

```{r Example Bivariate Map Theme Preview, fig.height = 5}
## Bivariate preview: suitability as fill (red = low, green = high) with
## prediction SE mapped to alpha, so uncertain pixels fade toward white
## rather than needing a second 3x3 legend grid.
df_biv_demo <- merge(rasterToDF(suit_r_demo), rasterToDF(se_r_demo), by = c('x', 'y'))
names(df_biv_demo)[3:4] <- c('value', 'se')

p_biv_preview <- mapPanelBivariate(df_biv_demo, label = 'Full domain (suitability x SE)', crs = demo_crs, speckle = speck_demo)
p_biv_preview
ggsave(file.path(fig_dir, 'bivariate_preview.png'), p_biv_preview, width = 7, height = 6, dpi = 300)
```
