---
title: "SDManuscript maps"
format: html
---

Builds the manuscript's map figures and writes each composite to `figures/`
as a PNG. Kept separate from SDManuscript.Rmd because the final rasters are
large and slow to crop/mask/render - knitting the manuscript itself
shouldn't have to pay that cost every time. Render this file whenever the
source rasters change; SDManuscript.Rmd just pulls in whatever PNGs it
finds in `figures/` via `knitr::include_graphics()`.

```{r setup}
library(ggplot2)
library(patchwork)
library(sf)
library(terra)
source('theme_manuscript.R')

fig_dir <- 'figures'
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

## Site label shown on every map panel/legend for the Crested Butte-area
## site - "Gothic" (the actual named locality, home of RMBL) rather than
## the town of Crested Butte itself. The internal object/list key stays
## 'gothic' throughout so it doesn't drift from this label again.
site_label_gothic <- 'Gothic'
```

## Final suitability and density maps


```{r zoom bboxes}
## Manually chosen zoom windows for the density figure's inset panels, read
## off a basemap in WGS84 lon/lat - each entry is NW corner then SE corner
## (lat, lon). Kept in WGS84 here; reprojected to the raster's CRS wherever
## they're actually used to crop (see `zoom_bbs` below).
zoom_corners <- list(
  gothic = c(nw_lat = 39.015644, nw_lon = -107.034565, se_lat = 39.007642, se_lon = -107.0207323),
  coche  = c(nw_lat = 38.177820, nw_lon = -106.705831, se_lat = 38.174130, se_lon = -106.696740),
  antero = c(nw_lat = 38.984090, nw_lon = -105.993880, se_lat = 38.974038, se_lon = -105.976990)
)

zoom_bbs_ll <- lapply(zoom_corners, function(z) {
  sf::st_bbox(
    c(xmin = z[['nw_lon']], xmax = z[['se_lon']], ymin = z[['se_lat']], ymax = z[['nw_lat']]),
    crs = 4326
  )
})
```

```{r Final Suitability and Density Maps, eval = FALSE}
## TODO: paths below point to the final refit outputs (2.8) - not yet
## produced (that model uses ground-truth data still being collected) -
## so this chunk stays eval=FALSE until they exist.

bbs_sf <- sf::st_read(
  file.path('..', '..', 'data', 'Cartography', 'bboxes_for_manuscript.gpkg'),
  quiet = TRUE
)
## 'coche' box was shifted 8km west and padded 2km N/S (see
## data/Cartography/bboxes_for_manuscript.gpkg) so its zoomed inset stops
## clipping the western edge of the surveyed area.
region_names <- c('gothic', 'coche', 'antero')
bbs <- setNames(lapply(1:3, function(i) sf::st_bbox(bbs_sf[i, ])), region_names)
## letter each region's own inset panel carries in the composite below -
## used to cross-reference that inset from its box on the full-domain panel
bbs_tags <- c(gothic = 'B', coche = 'C', antero = 'D')

suit_r <- terra::rast(file.path('..', '..', 'results', 'suitability_maps', 'final-Pr.tif'))
aoa_r  <- terra::rast(file.path('..', '..', 'results', 'suitability_maps', 'final-AOA.tif'))
dens_r <- terra::rast(file.path('..', '..', 'results', 'count_models', 'final-density.tif'))

## AOA == 0 pixels are outside the model's area of applicability - not real
## predictions - so null them out, and speckle that area on the plot rather
## than leaving it a blank gap.
suit_r  <- mask_aoa(suit_r, aoa_r)
speck   <- aoa_speckle(aoa_r)

## Binary threshold-cut outline (sens/spec_sens), drawn as a black line over
## the continuous suitability fill. `cost_thresh` in
## Second_groundtruth/DesignGroundTruth.R uses 'spec_sens' as the chosen
## binary cut for this project, so that's the layer used here.
## TODO: wire this up once a final-resolution thresholds raster (see
## Second_groundtruth/make_threshold_raster_masks.R for the pattern used at
## the search-iteration resolutions) exists alongside final-Pr.tif; no such
## raster exists yet for the final refit; needs the same care taken in
## transect_placement.R (dedupe identical patches, drop patches < 2 pixels)
## before treating its outline as meaningful.
## thresh_r      <- terra::rast(file.path('..', '..', 'results', 'suitability_maps', 'final-thresholds.tif'))
## cut_outline   <- sf::st_as_sf(terra::as.polygons(thresh_r[['spec_sens']] == 1, dissolve = TRUE))
## cut_outline   <- cut_outline[cut_outline[[1]] == 1, ]
## cut_line_layer <- geom_sf(data = cut_outline, inherit.aes = FALSE, fill = NA, color = 'black', linewidth = 0.4)

## ---- final habitat suitability: full domain (left, full height) + the
## three focal regions to the right, each panel the same size so smaller
## regions read at a visibly larger scale than the full domain. North arrow
## sits on the Gothic inset (not panel A) so it doesn't compete with the
## bbox/tag chrome already crowding the full-domain overview. ----
suit_full   <- mapPanel(
  rasterToDF(suit_r),
  scale_fill_suitability,
  'Full domain',
  bboxes = bbs,
  bbox_tags = bbs_tags,
  speckle = speck
  )
suit_gothic <- mapPanel(rasterToDF(suit_r, bbs$gothic), scale_fill_suitability, site_label_gothic, speckle = speck, north_arrow = TRUE)
suit_coche  <- mapPanel(rasterToDF(suit_r, bbs$coche),  scale_fill_suitability, 'Cochetopa', speckle = speck)
suit_antero <- mapPanel(rasterToDF(suit_r, bbs$antero), scale_fill_suitability, 'Antero', speckle = speck)

p_suit_final <- (suit_full + (suit_gothic / suit_coche / suit_antero) +
  plot_layout(widths = c(2, 1), guides = 'collect')) +
  plot_annotation(
    tag_levels = 'A',
    title = 'Final ENM for Eriogonum coloradense',
    subtitle = 'Post field work phase 1, and ground truth iteration') &
  theme_manuscript_tags() &
  theme_manuscript_legend() &
  theme(legend.position = 'bottom')
p_suit_final
ggsave(file.path(fig_dir, 'suitability_final.png'), p_suit_final, width = 9, height = 7, dpi = 300)
```

```{r Suitability Iteration Diffs Appendix, eval = FALSE}
## Appendix figure: iter0 (pre-existing occurrence records only) vs. iter1
## (refit after the first round of ground-truthing) suitability, and their
## difference, alongside the final (iter1 + boundary-transect data) map
## from the chunk above - so a reader can see how much each ground-truth
## round actually moved the prediction.
## TODO: no iteration-0 final-resolution suitability raster exists yet in
## results/suitability_maps (only a single 3-arc-second iter1 test run is
## available - see the 'Theme preview' section below); build this once
## iter0 and iter1 final-resolution rasters both exist, following the same
## mask_aoa()/rasterToDF()/mapPanel() pattern as the chunk above.
## iter0_r   <- terra::rast(file.path('..', '..', 'results', 'suitability_maps', 'iter0-Pr.tif'))
## iter1_r   <- suit_r  # from the chunk above
## diff_r    <- iter1_r - iter0_r
## diff_full <- mapPanel(rasterToDF(diff_r), scale_fill_diff, 'Iter1 - Iter0', bboxes = bbs, bbox_tags = bbs_tags)
```

```{r Final Density Maps, eval = FALSE}
## ---- final census/density surfaces: one row per site, stacked
## vertically (no full-domain panel here - density is only modelled/
## meaningful within the surveyed regions). Each row pairs the site-level
## map (with a nested zoom bbox outlined) against a further zoomed-in crop
## of that nested box. All sites share one fill scale/legend so density is
## comparable across sites, not just within one. ----
## Panels A, C, E (the main, non-zoomed column) reuse the exact same `bbs`
## object the suitability figure above crops with, so their extents match
## that figure's panels B/C/D by construction - don't recompute bboxes
## separately for this figure.
## Zoom windows are the manually chosen boxes from `zoom_bbs_ll` (WGS84),
## reprojected to the density raster's CRS.
dens_nested <- lapply(zoom_bbs_ll[names(bbs)], function(z) {
  sf::st_bbox(sf::st_transform(sf::st_as_sfc(z), terra::crs(dens_r)))
})

dens_df_gothic <- rasterToDF(dens_r, bbs$gothic)
dens_df_coche  <- rasterToDF(dens_r, bbs$coche)
dens_df_antero <- rasterToDF(dens_r, bbs$antero)
dens_limits <- range(c(dens_df_gothic$value, dens_df_coche$value, dens_df_antero$value), na.rm = TRUE)
dens_scale <- function() scale_fill_density(limits = dens_limits)

## TODO: panels B/D/F (the zoomed crops) are tight enough that a named
## sub-locality within each crop may be worth labelling on the map itself
## (distinct from the region-level 'Gothic'/'Cochetopa'/'Antero' label the
## main panel already carries) - needs the actual locality names, not
## guessed here.
dens_gothic_main <- mapPanel(dens_df_gothic, dens_scale, site_label_gothic, bboxes = list(zoom = dens_nested$gothic))
dens_coche_main  <- mapPanel(dens_df_coche,  dens_scale, 'Cochetopa', bboxes = list(zoom = dens_nested$coche))
dens_antero_main <- mapPanel(dens_df_antero, dens_scale, 'Antero', bboxes = list(zoom = dens_nested$antero))
## Zoom panels (B/D/F): a tight enough crop puts ggspatial's default scale-
## bar target (25% of panel width) right at its internal 1600m km-vs-m
## bucket boundary, which is what rendered panel F as '1000 m' instead of
## '1 km'. A larger width_hint pushes the target comfortably past that
## boundary so these consistently land on km.
dens_gothic_zoom <- mapPanel(rasterToDF(dens_r, dens_nested$gothic), dens_scale, paste(site_label_gothic, '(zoom)'), scale_width_hint = 0.35)
dens_coche_zoom  <- mapPanel(rasterToDF(dens_r, dens_nested$coche),  dens_scale, 'Cochetopa (zoom)', scale_width_hint = 0.35)
dens_antero_zoom <- mapPanel(rasterToDF(dens_r, dens_nested$antero), dens_scale, 'Antero (zoom)', scale_width_hint = 0.35)

p_dens_final <- (
  (dens_gothic_main | dens_gothic_zoom) /
  (dens_coche_main | dens_coche_zoom) /
  (dens_antero_main | dens_antero_zoom) +
  plot_layout(guides = 'collect')) +
  plot_annotation(
    tag_levels = 'A',
    title = 'Final Predicted Density, plants per pixel',
    subtitle = 'Post field work phase 1, and ground truth iteration') &
  theme_manuscript_tags() &
  theme_manuscript_legend() &
  theme(legend.position = 'bottom', legend.justification = 'center')
p_dens_final
ggsave(file.path(fig_dir, 'density_final.png'), p_dens_final, width = 8, height = 10, dpi = 300)
```

## Theme preview (real test raster, no final refit yet)

```{r Example Maps Theme Preview, fig.height = 6}
## Preview of theme_manuscript_map() styling using a real single-run
## suitability/AOA/SE triple dropped in results/suitability_maps for this
## purpose (not the final refit - that's still above, eval=FALSE until it
## exists). Delete this chunk once those are back to eval=TRUE.
## No real density test raster yet, so that panel stays on example_surface().
demo_fp <- file.path('..', '..', 'results', 'suitability_maps')
suit_r_demo <- terra::rast(file.path(demo_fp, '3arc-Iteration1-PA1:DO:0-Seed1017-Pr.tif'))
aoa_r_demo  <- terra::rast(file.path(demo_fp, '3arc-Iteration1-PA1:DO:0-Seed1017-AOA.tif'))
se_r_demo   <- terra::rast(file.path(demo_fp, '3arc-Iteration1-PA1:DO:0-Seed1017-SE.tif'))

## AOA == 0 pixels are outside the model's area of applicability - not real
## predictions - so null them out before plotting.
suit_r_demo <- mask_aoa(suit_r_demo, aoa_r_demo)
se_r_demo   <- mask_aoa(se_r_demo, aoa_r_demo)
demo_crs    <- terra::crs(suit_r_demo)
speck_demo  <- aoa_speckle(aoa_r_demo)

bbs_sf_demo <- sf::st_transform(
  sf::st_read(file.path('..', '..', 'data', 'Cartography', 'bboxes_for_manuscript.gpkg'), quiet = TRUE),
  demo_crs
)
region_names_demo <- c('gothic', 'coche', 'antero')
bbs_demo <- setNames(lapply(1:3, function(i) sf::st_bbox(bbs_sf_demo[i, ])), region_names_demo)
bbs_tags_demo <- c(gothic = 'B', coche = 'C', antero = 'D')
dens_r_demo <- example_surface(sf::st_bbox(bbs_sf_demo), res = 100, range = c(0, 40), seed = 28)

## ---- suitability: full domain (left) + the three focal regions to the
## right. North arrow sits on the Gothic inset, not the full-domain
## overview panel, so it doesn't collide with the bbox/tag chrome there. ----
suit_full_demo   <- mapPanel(rasterToDF(suit_r_demo), scale_fill_suitability,
                              'Full domain', crs = demo_crs, bboxes = bbs_demo, bbox_tags = bbs_tags_demo, speckle = speck_demo)
suit_gothic_demo <- mapPanel(rasterToDF(suit_r_demo, bbs_demo$gothic), scale_fill_suitability, site_label_gothic, crs = demo_crs, speckle = speck_demo, north_arrow = TRUE)
suit_coche_demo  <- mapPanel(rasterToDF(suit_r_demo, bbs_demo$coche),  scale_fill_suitability, 'Cochetopa', crs = demo_crs, speckle = speck_demo)
suit_antero_demo <- mapPanel(rasterToDF(suit_r_demo, bbs_demo$antero), scale_fill_suitability, 'Antero', crs = demo_crs, speckle = speck_demo)

p_suit_preview <- (suit_full_demo + (suit_gothic_demo / suit_coche_demo / suit_antero_demo) +
  plot_layout(widths = c(2, 1), guides = 'collect')) +
  plot_annotation(
    tag_levels = 'A',
    title = 'Final ENM for Eriogonum coloradense',
    subtitle = 'Post field work phase 1, and ground truth iteration') &
  theme_manuscript_tags() &
  theme_manuscript_legend() &
  theme(legend.position = 'bottom')
p_suit_preview
ggsave(file.path(fig_dir, 'suitability_preview.png'), p_suit_preview, width = 9, height = 7, dpi = 300)

## ---- density: one row per site, stacked vertically, each row pairing
## the site-level map (nested zoom bbox outlined) with a further zoomed-in
## crop of that nested box. All sites share one fill scale/legend. ----
dens_nested_demo <- lapply(bbs_demo, shrink_bbox, factor = 0.35)

dens_df_gothic_demo <- rasterToDF(dens_r_demo, bbs_demo$gothic)
dens_df_coche_demo  <- rasterToDF(dens_r_demo, bbs_demo$coche)
dens_df_antero_demo <- rasterToDF(dens_r_demo, bbs_demo$antero)
dens_limits_demo <- range(c(dens_df_gothic_demo$value, dens_df_coche_demo$value, dens_df_antero_demo$value), na.rm = TRUE)
dens_scale_demo <- function() scale_fill_density(limits = dens_limits_demo)

dens_gothic_main_demo <- mapPanel(dens_df_gothic_demo, dens_scale_demo, site_label_gothic, crs = demo_crs, bboxes = list(zoom = dens_nested_demo$gothic))
dens_coche_main_demo  <- mapPanel(dens_df_coche_demo,  dens_scale_demo, 'Cochetopa', crs = demo_crs, bboxes = list(zoom = dens_nested_demo$coche))
dens_antero_main_demo <- mapPanel(dens_df_antero_demo, dens_scale_demo, 'Antero', crs = demo_crs, bboxes = list(zoom = dens_nested_demo$antero))
## scale_width_hint bumped on the zoom panels - see the matching comment in
## the 'Final Density Maps' chunk above (ggspatial's m/km bucket boundary).
dens_gothic_zoom_demo <- mapPanel(rasterToDF(dens_r_demo, dens_nested_demo$gothic), dens_scale_demo, paste(site_label_gothic, '(zoom)'), crs = demo_crs, scale_width_hint = 0.35)
dens_coche_zoom_demo  <- mapPanel(rasterToDF(dens_r_demo, dens_nested_demo$coche),  dens_scale_demo, 'Cochetopa (zoom)', crs = demo_crs, scale_width_hint = 0.35)
dens_antero_zoom_demo <- mapPanel(rasterToDF(dens_r_demo, dens_nested_demo$antero), dens_scale_demo, 'Antero (zoom)', crs = demo_crs, scale_width_hint = 0.35)

p_dens_preview <- (
  (dens_gothic_main_demo | dens_gothic_zoom_demo) /
  (dens_coche_main_demo | dens_coche_zoom_demo) /
  (dens_antero_main_demo | dens_antero_zoom_demo) +
  plot_layout(guides = 'collect')) +
  plot_annotation(
    tag_levels = 'A',
    title = 'Final Predicted Density, plants per pixel',
    subtitle = 'Post field work phase 1, and ground truth iteration') &
  theme_manuscript_tags() &
  theme_manuscript_legend() &
  theme(legend.position = 'bottom', legend.justification = 'center')
p_dens_preview
ggsave(file.path(fig_dir, 'density_preview.png'), p_dens_preview, width = 8, height = 10, dpi = 300)
```

```{r Example Bivariate Map Theme Preview, fig.height = 5}
## Bivariate preview: suitability as fill (red = low, green = high) with
## prediction SE mapped to alpha, so uncertain pixels fade toward white
## rather than needing a second 3x3 legend grid.
df_biv_demo <- merge(rasterToDF(suit_r_demo), rasterToDF(se_r_demo), by = c('x', 'y'))
names(df_biv_demo)[3:4] <- c('value', 'se')

p_biv_preview <- mapPanelBivariate(df_biv_demo, label = 'Full domain (suitability x SE)', crs = demo_crs, speckle = speck_demo)
p_biv_preview
ggsave(file.path(fig_dir, 'bivariate_preview.png'), p_biv_preview, width = 7, height = 6, dpi = 300)
```
