---
title: "Evaluate Ground Truthing 2024"
author: "steppe"
output: html_document
---

We are interested in determining the accuracy of the existing geolocation of points. 
Many of these were revisited in 2024, and plants found as close to possible as them were recorded. 
If no plants could be founding within 50m or so the point was considered an absence at that time. 

```{r}
library(sf)
library(tidyverse)
library(terra)
set.seed(23)
```

```{r}
occ_data <- bind_rows(
  st_read('../data/collections/occurrences_coloradense/occurrences.shp', quiet = T) %>% 
    mutate(Occurrence = 1) %>% 
    st_transform(32613)
) 
```


I made a simple mistake when saving the pts to ground truth, I did not write out an ID from the initial process. 
Hopefully all of the ground verified points can be recovered by a combination of dates the observation was made, and their locations. 

```{r}
all_hist_pres <- st_read('../data/GroundTruthPts/Revisits-cp.shp') 

iter1gt <- st_read('../data/GroundTruthing/Iteration1Pts.shp', quiet = TRUE) |>
  filter(str_detect(Site, 'p.*')) |>
  mutate(
    Site = str_to_upper(Site)) |>
  left_join(st_drop_geometry(all_hist_pres), by = c('Site' = 'UID'))

table(sf::st_nearest_feature(occ_data, iter1gt)) 
# not super informative, because I didn't cover the full range of the species. 
# i'd wager the highest number point is simply my easternmost point!

table(sf::st_nearest_feature(iter1gt, occ_data)) 
# whereas this indicates most points I visited are close to a neighbor in the historic data set

table(occ_data$eventDate)[ table(occ_data$eventDate) > 1]
# yep lot's of repeat dates!!! hopefully from multiple observations... 
```

Points will need to be joined on both date and coordinates. 
While most records from the same day seem to be duplicates of a single observation, 
I am not 100% certain of this. 

```{r join points}
# neither method would be enough to match on on their own accord. 
toclean <- left_join(
  st_drop_geometry(all_hist_pres),
  filter(occ_data, Occurrence == 1), 
  by = 'eventDate') |>
  group_by(UID)

toclean %>% 
  group_by(UID) |>
  mutate(n = n()) |>
  filter(n == 1)

```