---
title: "Generate Absences for Modelling"
author: "steppe"
output: html_document
---

## Iteration 1
The goal of the model from iteration 1 is to generate many more presences which 
can be used for generating larger training/test data set. 

Accordingly at this stage we focus more on elucidating the currently known 
distribution of the species at the expense of expanding the expected distribution. 

```{r}
library(tidyverse)
library(sf)
library(terra)
source('functions.R')
set.seed(27)
```



First we will reduce the number of Presences in the data set to ensure that we do not have any in the same cells at 90m. 
Ten we will apply spThin to them, to avoid underestimating suitable habitat in less botanized areas. 
```{r}

occ <-  st_read('../data/collections/occurrences_coloradense/occurrences.shp', quiet = T) %>% 
  mutate(
    Longitude = unlist(map(.$geometry,1)),
    Latitude = unlist(map(.$geometry,2)), 
    Species = 'Eriogonum coloradense'
  ) %>% # do this to explictely drop repeat observations
  distinct(Longitude, Latitude, .keep_all = T)

thinned <- spThin::thin(
  loc.data = occ, lat.col = 'Latitude', long.col = 'Longitude', spec.col = 'Species',
  thin.par = 0.09, reps = 100, write.files = FALSE, write.log.file = FALSE, 
  locs.thinned.list.return = TRUE)

thinned <- thinned[[which.max(unlist(lapply(thinned, nrow)))]] %>% 
  st_as_sf(coords = c(x = 'Longitude', y = 'Latitude'), crs = 4326)

occ <- occ[ lengths(st_intersects(occ, thinned)) > 0, ] %>% 
  dplyr::select(-Longitude, -Latitude) %>% 
  st_transform(32613)

rm(thinned)
```

We will generate pseudo-absences which are from environments notable distinct from our presence records. 
```{r iteration 1 absence generation, eval = F}

p2proc <- '../data/spatial/processed'
arc3 <- rastReader('dem_3arc', p2proc) 

# we want to be SURE that we are finding more populations, or segments of populations
# during this iteration of field sampling. Accordingly, we will want a model which 
# is more conservative in predicting suitable habitat. 

PA_possible <- sdm::background(arc3, 
                n = nrow(occ) + (nrow(occ)*0.1),
                method = 'eDist', sp = occ) %>% 
  select(x, y) %>% 
  st_as_sf(coords = c(x = 'x', y = 'y'), crs = 32613) %>% 
  mutate(Occurrence = 0, ID = 1:nrow(.))

# points were then manually investigated, and those which could foresee-ably be within
# subjectively evaluated suitable habitat were removed. 

st_write(PA_possible, 
         '../data/spatial/raw/pseudo-absences/iteration1/absences.shp', append = F)

# these points were identified in QGIS as likely occurring in suitable habitat. 
removals <- c(19, 41, 42, 46, 67, 76)
PA_possible <- filter(PA_possible, ! ID %in% removals)

if(nrow(PA_possible) > nrow(occ)){
  PA_possible <- PA_possible[sample(1:nrow(PA_possible), size = nrow(occ)),]
}

st_write(PA_possible, append = F,
         '../data/spatial/processed/pseudo-absences/iteration1/absences.shp')
```

