---
title: "Modelling Eriogonum coloradense Presence and Abundance"
author: "steppe"
output: html_document
---

```{r, warning = F, message = F}
library(sf)
library(tidyverse)
library(terra)
library(caret)
library(bonsai)
library(parsnip)
library(ranger)
library(lightgbm)
library(future)
library(dials)
library(tune)
library(finetune)
source('functions.R')
set.seed(23)



```

```{r}
p <- '../data/Data4modelling'
f <- file.path(p, list.files(p))
```

Model presence for iteration 0, that is historic records only. 

Model presence for iteration 1, both the historic and 2024 ground truth data. 

```{r Run all models, eval = F}

abs <- st_read('../data/Data4modelling/iter1-pa.gpkg')
m30 <- sf::st_read("../data/Data4modelling/10m-presence-iter1.gpkg") %>% 
  rename(Occurrence = Presenc) |>
  sf::st_as_sf() 
### we know that 1:1 absence to presence is far too low when including the 'local'
# absences. 
m30 <- bind_rows(m30, abs)
m30 <- filter(m30, st_is(m30, "POINT")) |>
  select(Occurrence)

distOrder_PAratio_simulator( # this does all the presence absence modelling. = bad function name
  x = m30, distOrder = 0, PAratio = 3.0,
  resolution = 10, iteration = 1, se_prediction = FALSE,
  train_split = 0.9, p2proc = '../data/spatial/processed'
  )

```

# playing with count prediction 


```{r Partition data and clean for modelling}

p <- '/media/steppe/hdd/EriogonumColoradenseTaxonomy/results/suitability_maps'
f <- list.files(p, pattern = '-Pr.tif')

lapply(f[8:11], wrapper)
```


Modelling count data has not been very successful, nor did the field work really develop any strong insights on what drives the abundance of the plants. 
Rather the abundance seemed random within populations, and between - with the exception of Cocheotopa where plants were much smaller and incredibly more abundant than elsewhere. 

The null hypothesis for estimating population sizes will be the arithmetic mean per populations. 

The use of kriging interpolation, where we simple smooth averages out between the plots, will be our null hypothesis for any spatial estimates of population size. 

```{r}
library(iml)
library(xgboost) 
d_w_vars <- wrapper('1-3arc-Iteration1-PA1:2.7DO:0-Pr.tif')
```



```{r}
mod <- readRDS('../results/CountModels/models/1-3arc-Iteration1-PA1:2.7-tweedie.rds')$Model
vars <- as.character(attr(mod[["preproc"]][["terms"]], 'variables')) # fns will run on all 
vars <- vars[2:length(vars)] # vars even if not in model - but produce null wts as expected. 

test_indices <- as.numeric(
  read.delim(
    '../results/CountModels/test_data/twin_indx-1-3arc.txt', header = F, sep = ' ')
  )

test_dat <- d_w_vars[test_indices,] |>
  select(any_of(vars)) |>
  st_drop_geometry()

predict.fun <- function(object, newdata){
  newData_x = xgb.DMatrix(data.matrix(newdata), missing = NA)
  results<-predict(mod, newData_x)
}

subby <- test[which(names(test) != "Prsnc_All")]
predictor <- Predictor$new(
  model = mod,
  data = subby, 
  y = test_dat$Prsnc_All, 
  predict.function = predict.fun
  )

# this does not work... 
shapley <- Shapley$new(predictor, x.interest = test) 

private$sampler$feature.names

# these both run   
int_out <- Interaction$new(predictor, grid.size = 10)
plot(int_out)

prsuit_int <- Interaction$new(predictor, feature = "Pr.SuitHab", grid.size = 15)
plot(prsuit_int)

######## try to get shap with a different package. 
library(kernelshap)
library(shapviz)

s <- kernelshap::kernelshap(mod, X = test_dat) 
sv <- shapviz::shapviz(s)
shapviz::sv_importance(sv, kind = "bee") +
  theme_bw()

shapviz::sv_importance(sv, kind = "bar") +
  theme_bw()

shapviz::sv_dependence(sv, v = colnames(test_dat[, -1]))

shapviz::sv_waterfall(sv) +
  theme(axis.text = element_text(size = 11))
```
