---
title: "Modelling Eriogonum coloradense Presence and Abundance"
author: "steppe"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/media/steppe/hdd/EriogonumColoradenseTaxonomy/scripts')
```

```{r, warning = F, message = F}
library(sf)
library(tidyverse)
library(terra)
library(caret)
library(bonsai)
library(parsnip)
library(ranger)
library(lightgbm)
library(future)
library(dials)
library(tune)
library(finetune)
source('functions.R')
set.seed(23)
```

```{r}
p <- '../data/Data4modelling'
f <- file.path(p, list.files(p))
```

Model presence for iteration 0, that is historic records only. 

Model presence for iteration 1, both the historic and 2024 ground truth data. 

```{r Run all models, eval = F}

abs <- st_read('../data/Data4modelling/iter1-pa.gpkg')
m30 <- sf::st_read("../data/Data4modelling/10m-presence-iter1.gpkg") %>% 
  rename(Occurrence = Presenc) |>
  sf::st_as_sf() 
### we know that 1:1 absence to presence is far too low when including the 'local'
# absences. 
m30 <- bind_rows(m30, abs)
m30 <- filter(m30, st_is(m30, "POINT")) |>
  select(Occurrence)

distOrder_PAratio_simulator( # this does all the presence absence modelling. = bad function name
  x = m30, distOrder = 0, PAratio = 3.0,
  resolution = 10, iteration = 1, se_prediction = FALSE,
  train_split = 0.9, p2proc = '../data/spatial/processed'
  )


rastReader <- function(x, p2proc){

  # create paths for the first level of files. 
  f <- file.path(p2proc, x, list.files(file.path(p2proc, x), pattern = '.tif$'))

  # create paths for the geomorphology files.
  geo_f <- file.path(p2proc, x, 'geomorphology',
                     list.files(file.path(p2proc, x, 'geomorphology')))
  
  geo_f <- geo_f[! grepl('D8pntr|basins|DEM', geo_f) ] # remove the D8pntr & Basins 
  
  # combine the file paths
  f <- c(f, geo_f)
  s <- terra::rast(f)
  
}

d <- rastReader('dem_1arc',  '../data/spatial/processed')
write.csv(
  data.frame('rast_lyr_name' = names(d)), 
  '../data/spatial/processed/raster_lyr_names.csv'
  )

```

# playing with count prediction 

```{r Partition data and clean for modelling, eval = F}

p <- '../results/suitability_maps'
f <- list.files(p, pattern = '-Pr.tif')

lapply(f[11], wrapper)
```


Modelling count data has not been very successful, nor did the field work really develop any strong insights on what drives the abundance of the plants. 
Rather the abundance seemed random within populations, and between - with the exception of Cocheotopa where plants were much smaller and incredibly more abundant than elsewhere. 

The null hypothesis for estimating population sizes will be the arithmetic mean per populations. 

The use of kriging interpolation, where we simple smooth averages out between the plots, will be our null hypothesis for any spatial estimates of population size. 

```{r}
library(iml)
library(xgboost) 
d_w_vars <- wrapper('3arc-Iteration1-PA1_3.3DO_0-Pr.tif')
```


```{r}
mod <- readRDS('../results/count_models/models/1-3arc-Iteration1-PA1_2.7-lgbm-poisson_spat.rds')$Model
vars <- as.character(attr(mod[["preproc"]][["terms"]], 'variables')) # fns will run on all 
vars <- vars[2:length(vars)] # vars even if not in model - but produce null wts as expected. 

test_indices <- as.numeric(
  read.delim(
    '../results/count_models/test_data/twin_indx-3arc.txt', header = F, sep = ' ')
  )

test_dat <- d_w_vars[test_indices,] |>
  select(any_of(vars)) |>
  st_drop_geometry()

predict.fun <- function(object, newdata){
  newData_x = xgb.DMatrix(data.matrix(newdata), missing = NA)
  results<-predict(mod, newData_x)
}

subby <- test_dat[which(names(test_dat) != "Prsnc_All")]
predictor <- Predictor$new(
  model = mod,
  data = subby, 
  y = test_dat$Prsnc_All, 
  predict.function = predict.fun
  )

```

```{r}
library(DALEX)
library(DALEXtra)
train_dat <- d_w_vars[-test_indices,]

explainer <- explain_tidymodels(
  mod,
  data = train_dat %>% select(-Prsnc_All) |> sf::st_drop_geometry(),
  y = train_dat$Prsnc_All,
  label = "tidy LightGBM"
)

pdp_ice <- model_profile(
  explainer, 
  variables = "Pr.SuitHab", 
  N = 50,              # sample 50 ICE curves
  center = TRUE,
  type = 'partial',
  grid_points = 40
) 


## we can compare models for a resolution using this chart - I am still not entirely
# sure what this means. 
resids <- model_performance(explainer)
plot(resids)
# create comparison plot of residuals for each model
# p1 <- plot(resids_glm, resids_rf, resids_gbm)
# p2 <- plot(resids_glm, resids_rf, resids_gbm, geom = "boxplot")
# gridExtra::grid.arrange(p1, p2, nrow = 1)

###
vi <- DALEX::variable_importance(explainer)
plot(vi)


df_pdp <- pdp_ice$agr_profiles |>
  janitor::clean_names() 

df_ice <- pdp_ice$cp_profiles |>
  janitor::clean_names()

ggplot(df_pdp, aes(x = x, y = yhat, group = ids)) +
  geom_line(data = df_ice, aes(x = pr_suit_hab), alpha = 0.4, color = "grey") +
  geom_line(color = "steelblue", size = 1) + # pdp curve+
  geom_rug(data = train_dat, aes(x = Pr.SuitHab), 
           sides = "b", alpha = 0.3, inherit.aes = F) +
  labs(
    x = "Probability of Suitable Habitat",
    y = "Predicted Plants per Cell",
    title = "Number of plants predicted per raster cell"
  ) +
  theme_minimal()
  
```

Accumulated local effects plots

```{r}

ale_prof <- model_profile(
  explainer,                # your DALEX explainer
  variables   = "Pr.SuitHab",
  type        = "accumulated",  # <-- use ALE here
  N           = 100,            # number of observations to sample
  grid_points = 40,             # number of bins to split variable
  center      = TRUE            # center effect at zero
)

plot(ale_prof) +
  labs(
    title = "Accumulated Local Effects (ALE) for Pr.SuitHab",
    x     = "Probability of Suitable Habitat",
    y     = "ALE"
  ) +
  theme_minimal()

```

