---
title: "Modelling Eriogonum coloradense Presence and Abundance - Iteration 0"
author: "steppe"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/media/steppe/hdd/EriogonumColoradenseTaxonomy/scripts')
```

```{r, warning = F, message = F}
library(sf)
library(tidyverse)
library(terra)
library(caret)
library(ranger)
library(future)
source('../functions.R')
```

Model presence at each of three spatial resolutions (3 arc-second/90m, 1 arc-second/30m,
1/3 arc-second/10m) for iteration 0 (historic records only, pre-2024 ground-truthing),
searching the presence:absence ratio adaptively at each resolution rather than fixing it -
see `adaptive_PAratio_search()` in functions.R. The absence pool is shared with iteration 1
(`iter1-pa.gpkg`) since no iteration-0-specific pool exists.

```{r Shared absence pool}
abs <- st_read('../../data/Data4modelling/iter1-pa.gpkg')
areas <- st_read('../../data/GroundTruthPts/ground_truth_median-VISITED.gpkg', layer = 'areas')
```

## 3 arc-second (90m)

```{r Resolution 90m, eval = F}
m90 <- sf::st_read('../../data/Data4modelling/90m-presence-iter0.gpkg') %>%
  rename(Occurrence = Presenc) |>
  sf::st_as_sf()
m90 <- bind_rows(m90, abs)
m90 <- filter(m90, st_is(m90, "POINT")) |>
  select(Occurrence)

search_90 <- adaptive_PAratio_search(
  x = m90, distOrder = 0, resolution = 90, iteration = 0,
  train_split = 0.8, p2proc = '../../data/spatial/processed'
  )

search_90_classic <- adaptive_PAratio_search(
  x = m90, distOrder = 0, resolution = 90, iteration = 0, train_split = 0.8,
  p2proc = '../../data/spatial/processed',
  spatial_split = FALSE,
  results_root = file.path(PROJ_ROOT, 'results_classicsplit')
)

lopo_90 <- population_lopo_search(
  x = m90, distOrder = 0, resolution = 90, iteration = 0,
  PAratio = search_90$best_ratio, areas = areas,
  p2proc = '../../data/spatial/processed',
  results_root = file.path(PROJ_ROOT, 'results_populationsplit')
)
```

## 1 arc-second (30m)

```{r Resolution 30m, eval = F}
m30 <- sf::st_read('../../data/Data4modelling/30m-presence-iter0.gpkg') %>%
  rename(Occurrence = Presenc) |>
  sf::st_as_sf()
m30 <- bind_rows(m30, abs)
m30 <- filter(m30, st_is(m30, "POINT")) |>
  select(Occurrence)

search_30 <- adaptive_PAratio_search(
  x = m30, distOrder = 0, resolution = 30, iteration = 0,
  train_split = 0.8, p2proc = '../../data/spatial/processed'
  )

search_30_classic <- adaptive_PAratio_search(
  x = m30, distOrder = 0, resolution = 30, iteration = 0, train_split = 0.8,
  p2proc = '../../data/spatial/processed',
  spatial_split = FALSE,
  results_root = file.path(PROJ_ROOT, 'results_classicsplit')
)

lopo_30 <- population_lopo_search(
  x = m30, distOrder = 0, resolution = 30, iteration = 0,
  PAratio = search_30$best_ratio, areas = areas,
  p2proc = '../../data/spatial/processed',
  results_root = file.path(PROJ_ROOT, 'results_populationsplit')
)
```

## 1/3 arc-second (10m)

```{r Resolution 10m, eval = F}
m10 <- sf::st_read('../../data/Data4modelling/3m-presence-iter0.gpkg') %>%
  rename(Occurrence = Presenc) |>
  sf::st_as_sf()
m10 <- bind_rows(m10, abs)
m10 <- filter(m10, st_is(m10, "POINT")) |>
  select(Occurrence)

search_10 <- adaptive_PAratio_search(
  x = m10, distOrder = 0, resolution = 10, iteration = 0,
  train_split = 0.8, p2proc = '../../data/spatial/processed'
  )

search_10_classic <- adaptive_PAratio_search(
  x = m10, distOrder = 0, resolution = 10, iteration = 0, train_split = 0.8,
  p2proc = '../../data/spatial/processed',
  spatial_split = FALSE,
  results_root = file.path(PROJ_ROOT, 'results_classicsplit')
)

lopo_10 <- population_lopo_search(
  x = m10, distOrder = 0, resolution = 10, iteration = 0,
  PAratio = search_10$best_ratio, areas = areas,
  p2proc = '../../data/spatial/processed',
  results_root = file.path(PROJ_ROOT, 'results_populationsplit')
)
```
