---
title: "Interpreting the Eriogonum coloradense Count/Density Model"
author: "steppe"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/media/steppe/hdd/EriogonumColoradenseTaxonomy/scripts')
```

```{r, warning = F, message = F}
library(sf)
library(tidyverse)
library(terra)
library(bonsai)
library(parsnip)
library(lightgbm)
library(xgboost)
library(iml)
library(DALEX)
library(DALEXtra)
library(janitor)
source('functions.R')
```

Modelling count data has not been very successful, nor did the field work really develop any strong insights on what drives the abundance of the plants.
Rather the abundance seemed random within populations, and between - with the exception of Cochetopa where plants were much smaller and incredibly more abundant than elsewhere.

The null hypothesis for estimating population sizes will be the arithmetic mean per population.

The use of kriging interpolation, where we simply smooth averages out between the plots, will be our null hypothesis for any spatial estimates of population size.

```{r Load suitability predictions and fitted count model, eval = F}
d_w_vars <- wrapper('3arc-Iteration1-PA1_3.3DO_0-Pr.tif')

mod <- readRDS('../results/count_models/models/1-3arc-Iteration1-PA1_2.7-lgbm-poisson_spat.rds')$Model
vars <- as.character(attr(mod[["preproc"]][["terms"]], 'variables')) # fns will run on all
vars <- vars[2:length(vars)] # vars even if not in model - but produce null wts as expected.

test_indices <- as.numeric(
  read.delim(
    '../results/count_models/test_data/twin_indx-3arc.txt', header = F, sep = ' ')
  )

test_dat <- d_w_vars[test_indices,] |>
  select(any_of(vars)) |>
  st_drop_geometry()

predict.fun <- function(object, newdata){
  newData_x = xgb.DMatrix(data.matrix(newdata), missing = NA)
  results<-predict(mod, newData_x)
}

subby <- test_dat[which(names(test_dat) != "Prsnc_All")]
predictor <- Predictor$new(
  model = mod,
  data = subby,
  y = test_dat$Prsnc_All,
  predict.function = predict.fun
  )
```

```{r Build DALEX explainer, eval = F}
train_dat <- d_w_vars[-test_indices,]

explainer <- explain_tidymodels(
  mod,
  data = train_dat %>% select(-Prsnc_All) |> sf::st_drop_geometry(),
  y = train_dat$Prsnc_All,
  label = "tidy LightGBM"
)

pdp_ice <- model_profile(
  explainer,
  variables = "Pr.SuitHab",
  N = 50,              # sample 50 ICE curves
  center = TRUE,
  type = 'partial',
  grid_points = 40
)

## we can compare models for a resolution using this chart - I am still not entirely
# sure what this means.
resids <- model_performance(explainer)
plot(resids)
# create comparison plot of residuals for each model
# p1 <- plot(resids_glm, resids_rf, resids_gbm)
# p2 <- plot(resids_glm, resids_rf, resids_gbm, geom = "boxplot")
# gridExtra::grid.arrange(p1, p2, nrow = 1)

###
vi <- DALEX::variable_importance(explainer)
plot(vi)


df_pdp <- pdp_ice$agr_profiles |>
  janitor::clean_names()

df_ice <- pdp_ice$cp_profiles |>
  janitor::clean_names()

ggplot(df_pdp, aes(x = x, y = yhat, group = ids)) +
  geom_line(data = df_ice, aes(x = pr_suit_hab), alpha = 0.4, color = "grey") +
  geom_line(color = "steelblue", size = 1) + # pdp curve+
  geom_rug(data = train_dat, aes(x = Pr.SuitHab),
           sides = "b", alpha = 0.3, inherit.aes = F) +
  labs(
    x = "Probability of Suitable Habitat",
    y = "Predicted Plants per Cell",
    title = "Number of plants predicted per raster cell"
  ) +
  theme_minimal()
```

Accumulated local effects plots

```{r Accumulated local effects, eval = F}
ale_prof <- model_profile(
  explainer,                # your DALEX explainer
  variables   = "Pr.SuitHab",
  type        = "accumulated",  # <-- use ALE here
  N           = 100,            # number of observations to sample
  grid_points = 40,             # number of bins to split variable
  center      = TRUE            # center effect at zero
)

plot(ale_prof) +
  labs(
    title = "Accumulated Local Effects (ALE) for Pr.SuitHab",
    x     = "Probability of Suitable Habitat",
    y     = "ALE"
  ) +
  theme_minimal()
```
